Robert C. Dunnell’s Systematics in Prehistory at 50

### 2021-06-04 analysis/paper/paper.Rmd
Further update

The notion of major paradigm shifts in the history of archaeology is characterized by a great deal of hyperbole and rhetorical maneuvering. That said, the late 1960s and 1970s really do stand out as revolutionary. In the US, the Binfordian steamroller increasingly hammered away at its culture-historical nemesis, while in the UK, David Clarke's formidable *Analytical Archaeology* was shaking up the establishment [@clarke1968a]. The reception and impact of these works has seen a great deal of attention in later years, especially as many researchers ar rediscovering the merits of Clarke's approach in particular [@nicholas2012][@lycett2018a]. 2021 marks the fiftieth anniversary of the publication of Robert C. Dunnell's (1971) diminutive yet dense *Systematics in Prehistory* - volume that is concerned with nothing but classification in archaeology. Published at the height of the debate between Culture History and New Archaeology, Dunnell's work was no less iconoclastic than that of Binford or Clarke, but it was considerably narrower in its endeavor to address a more fundamental issue that was and still is relevant to all branches of prehistoric archaeology, and especially to the study of the Paleolithic: systematics.

Dunnell himself was notorious and controversial, and his *Systematics* was received with mixed reactions by his contemporaries who commented - most negatively - both on his style of writing and the book's content [@spaulding1974][@tuggle1974][@bayard1973][@shenkel1973]. Like other precocious works of that tumultuous time *Systematics in Prehistory* today remains absent from most course reading lists and gathers dust on library shelves. As Lyman has recently shown, systematics takes up little space in contemporary archaeological research or teaching [@lyman2021]. If at least some of the success of Clarke and Binford can be attributed to them boldly tackling exciting and large-scale topics such as migration and adaptation using then avant-garde terminology, then corresponding obscurity of Dunnell's _Systematics_ can perhaps be attributed to him focusing on an issue that simply seemed to state. Yet, Dunnell was correct in his axiomatic insistence that rigorous classification comes before any other analysis or interpretation. Without consistent and explicit classification, any scientific discipline will inevitably fail to produce cumulative insights.

@@

@@

For Dunnell himself, the soul-searching that began with writing *Systematics in Prehistory* led him to discover evolutionary theory. In a series of subsequent papers, he forcefully argued for the benefits of a scientific and Darwinian archaeology @dunnell1980; @dunnell1982[]. While not making many friends, these writings have since become foundational for the development of evolutionary archaeology. Initially, following the lead of Dunnell, this approach was rather narrowly selectionist - treating artefacts as the hard parts of the human phenotype and selection acting on these as the main driver of change @obrien1990 - but which since has become more fully aligned with cultural evolutionary thinking in the form of dual-inheritance and niche construction theory [@riede2019][@marwick2006].

>>>>

The notion of major paradigm shifts in the history of archaeology is characterized by a great deal of hyperbole and rhetorical maneuvering. That said, the late 1960s and 1970s really do stand out as rather revolutionary. In the US, the Binfordian steamroller increasingly hammered away at its culture-historical nemesis, while in the UK, David Clarke's formidable *Analytical Archaeology* was shaking up the establishment with novel concepts, novel methods and an incisive rhetoric [@clarke1968a]. The reception and impact of these works has seen a great deal of attention in later years, especially as many researchers ar rediscovering the merits of Clarke's approach in particular [@nicholas2012][@lycett2018a]. 2021 marks the fiftieth anniversary of the publication of Robert C. Dunnell's (1971) diminutive yet dense *Systematics in Prehistory*, a volume that is concerned with nothing but classification in archaeology. Published at the height of the debate between Culture History and New Archaeology, Dunnell's work was no less iconoclastic than that of Binford or Clarke, but it was considerably narrower in its endeavor to address a more fundamental issue that was and still is relevant to all branches of prehistoric archaeology, and especially to the study of the Paleolithic: systematics.

Dunnell himself was notorious and controversial, and his *Systematics* was received with mixed reactions by his contemporaries who commented - mostly negatively - both on his style of writing and the book's content [@spaulding1974][@tuggle1974][@bayard1973][@shenkel1973]. Like other precocious works of that tumultuous time *Systematics in Prehistory* today remains absent from most course reading lists and gathers dust on library shelves. As Lyman has recently shown, systematics takes up little space in contemporary archaeological research or teaching [@lyman2021]. If at least some of the success of Clarke and Binford can be attributed to them boldly tackling exciting and large-scale topics such as migration and adaptation using then avant-garde terminology, then the corresponding obscurity [figure of citation frequency here?] of Dunnell's *Systematics* can perhaps be attributed to him focusing on an issue that simply seemed too uninteresting. Yet, Dunnell was correct in his axiomatic insistence that rigorous classification comes before any other analysis or interpretation. Without consistent and explicit classification, any scientific discipline will inevitably fail to produce cumulative insights.

@@

While Dunnell worked exclusively on the Holocene prehistory of the Americas, the ideosyncracies of archaeological classification are nowhere more apparent and acute than in Paleolithic archaeology. Rooted in French antiquarianism, the development of Palaeolithic systematics has been likened to 'accidents of history' @clark2009, and there are few periods or regions of the Palaeolithic that have not seen debate about the validity or otherwise of their analytical units @reynolds2019.

@@

For Dunnell himself, the soul-searching that began with writing *Systematics in Prehistory* led him to discover evolutionary theory. In a series of subsequent papers, he forcefully argued for the benefits of a scientific and Darwinian archaeology @dunnell1980; @dunnell1982. While not making many friends, these writings have since become foundational for the development of evolutionary archaeology. Initially, following the lead of Dunnell, this approach was rather narrowly selectionist - treating artefacts as the hard parts of the human phenotype and selection acting on these as the main driver of change @obrien1990 - but which since has become more fully aligned with cultural evolutionary thinking in the form of dual-inheritance and niche construction theory [@riede2019][@marwick2006].
